---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# climateChangeInAustralia

<!-- badges: start -->
![GitHub R package version](https://img.shields.io/github/r-package/v/shaunson26/climateChangeInAustralia)
[![R-CMD-check](https://github.com/Shaunson26/climateChangeInAustralia/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Shaunson26/climateChangeInAustralia/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of climateChangeInAustralia is to create an API to [Climate Change in Australia](https://www.climatechangeinaustralia.gov.au/) datasets.

- Links to dataset URLs
- Functions to download data subsets (NetCDF subsets)
- Functions and objects to help with data selection

```
# install.packages("devtools")
devtools::install_github("Shaunson26/climateChangeInAustralia")
```

```
library(climateChangeInAustralia)
```

```{r, include = F}
devtools::load_all()
```

## Source data

> https://www.climatechangeinaustralia.gov.au/ - Climate information, projections, tools and data.

They provide [datasets](https://www.climatechangeinaustralia.gov.au/en/obtain-data/download-datasets/) from various [models](https://www.climatechangeinaustralia.gov.au/en/overview/methodology/list-models/) and for 
a number of [variables](https://www.climatechangeinaustralia.gov.au/en/obtain-data/download-datasets/ccia-file-naming-convention/)
that include

1. Projected change (compared with 1986-2005) data ... +/- values
2. Plausible future climate data ... predicted values
3. Threshold datasets ... number of days above/below extremes e.g. >30, >35, >45

Data is accessed via navigation through their web directories, and the file names have
a [naming convention](https://www.climatechangeinaustralia.gov.au/en/obtain-data/download-datasets/ccia-file-naming-convention/)
describing (for the purpose of this package): variable, model, representative concentration pathway (RCP, greenhouse gas scenarios) and time period among other things.

> We crawled through their directories and created a JSON file with the links to all datasets. This can be
imported as a named nested list for ease of navigation to particular datasets using `ccia_dataset_urls()`

```{r}
ccia_datasets <- ccia_dataset_urls()

# dataset$variable_1$variable_1_1$datset_name
ccia_datasets$
  projected_change_data$
  Drought$
  `freq-duration`$
  `spi_Amon_ACCESS1-0_historical_r1i1p1_anntot-events_native.nc`
```

From there, there are various options to download the data e.g. HTTP, NetcdfSubset, among others

> Our download functions allow for the 8 'access' types by we will only going to deal with NetcdfSubset and HTTP from here on.

The NetcdfSubset access uses a REST API - `dataset/url?query=parameters`. There is
a dataset description XML that will be used to aid in data queries to avoid HTTP bad 
requests (ensuring we make valid queries. nSee examples later on...

### Future gridded change data (native grid)

"These data provide the 20-year averaged monthly, seasonal and annual changes for four 
time periods centred on 2030, 2050, 2070 & 2090 using a time-slice method."

- Variables include

```{r}
names(ccia_datasets$projected_change_data) |>
  cat(sep = '\n')
```

- 15 year range brackets ... but a single time range value that the data was centred on e.g. "2030-01-01"

```{r}
names(ccia_datasets$projected_change_data$Mean_Temperature) |>
  cat(sep = '\n')
```

- Multiple 'seasonal' variables within time range value ... 'annual', 'april', 'djf' (months of summer)
- small data files ... 0.25 to 10 Mb 

### Application-ready datasets (5km grids)

"Application-ready datasets "produce plausible future climate data at daily, monthly, 
seasonal and annual time steps on the high-resolution 5 km grid."

#### Daily time-seris

- Variables include

```{r}
names(ccia_datasets$application_ready_daily) |>
  cat(sep = '\n')
```

- 30 year range brackets ... with daily data available

```{r}
names(ccia_datasets$application_ready_daily$Mean_Temperature) |>
  cat(sep = '\n')
```

- Usually a single variable within the time range value
- **HUGE** data files 25 Gb ... and the servers often timeout ... expect to be doing multiple time slicing for these!

#### Monthly, seasonal, annual time-seris

- Variables include

```{r}
names(ccia_datasets$application_ready_aggregated) |>
  cat(sep = '\n')
```

- 30 year range brackets ... but often a single time range value e.g. 2030-01-01

```{r}
names(ccia_datasets$application_ready_aggregated$Mean_Temperature) |>
  cat(sep = '\n')
```

- Multiple 'seasonal' variables within time range value ... 'annual', 'april', 'djf' (months of summer)
- small data files ... 0.25 to 10 Mb 

### Threshold datasets (5km grids)

"These datasets are derived from the gridded application-ready daily time-series described above. 
The data have been analysed to find the number of days (on average) when ... max > 30-45, min > 20-30, min < 0-22"

- Variables include

```{r}
names(ccia_datasets$application_ready_threshold) |>
  cat(sep = '\n')
```

- 30 year range brackets ... 

```{r}
names(ccia_datasets$application_ready_threshold$Maximum_Temperature) |>
  cat(sep = '\n')
```

- `mon` monthly centered (time dimension) for 2 variables (>34, >35)
- `clim` annual, monthly, seasonal (as variables) .. possibly year range centered 

```{r}
# These can get a bit confusing
names(ccia_datasets$application_ready_threshold$Maximum_Temperature$`2016-2045`) |>
  cat(sep = '\n')
```

- Multiple 'seasonal' variables within `clim` time range value ... 'annual', 'april', 'djf' (months of summer)
- **Big** data files ... 700 Mb - 2 Gb ... expect to slice times

## Example use

With the dataset URL list

```{r}
ccia_datasets <- ccia_dataset_urls()
```

Navigate the list object to a file name (ends in `.nc`). This returns a relative URL

```{r}
# ara = application ready aggregated
ara_mean_temp_2016_access1_rcp45 <- 
  ccia_datasets$
  application_ready_aggregated$
  Mean_Temperature$
  `2016-2045`$
  `tas_aus_ACCESS1-0_rcp45_r1i1p1_CSIRO-MnCh-wrt-1986-2005-Scl_v1_mon_seasavg_2016-2045_clim.nc`

ara_mean_temp_2016_access1_rcp45
```

We then create a `httr2` request object using `ccia_create_dataset_request_url()`, and
that we will build upon. There are different access types to the 
file (they have different URL paths). In this package, we use `"NetcdfSubset"` and the
following functions expect an intial request object build with this. We _can_ build other
access URLs - `ccia_access_types` lists the 8 access types available.

```{r}
ara_mean_temp_2016_access1_rcp45_req <- 
  ccia_create_dataset_request_url(dataset_relative_url = ara_mean_temp_2016_access1_rcp45, 
                             access_type = ccia_access_types$NetcdfSubset)

ara_mean_temp_2016_access1_rcp45_req
```

Now we have the URL to the `"NetcdfSubset"` dataset but we need to append some
queries to subset the data. Generally we need to add queries for

  - the dataset variable(s)
  - the time span
  - lat/lon

The datasets can have few or many 'internal' variables (the 'names' of the columns
holding the data values, and whether there are multiple 'columns' of these). 
**We have the option of using `all` for all** variables - which is usually safe for 
small data calls, or we can can investigate the dataset using 
it's "dataset description" - an XML we download from a complete URL (minus query parameters).
This is achieved using `ccia_get_dataset_netcdf_subset_info()`

```{r}
ara_mean_temp_2016_access1_rcp45_info <-
  ccia_get_dataset_netcdf_subset_info(ara_mean_temp_2016_access1_rcp45_req)

ara_mean_temp_2016_access1_rcp45_info
```

The `vars$name` represents the 'columns' of the data, and the `TimeSpan` the time
ranges (some times there are strange time points - monthly centered, year range centered). 
Obtaining the dataset description will help in avoiding a bad request, however, the
server will return useful messages (these are printed to the console when it happens).
We use the data here to create a query to add to the request.

Requests use a REST API structure `dataset/url?query=parameter` and the function
`ccia_add_netcdf_subset_query()` helps here. In this example, we pick the 4 '3 month periods'
(seasons in Australia), use the single time value for this data, and a bounding box of coordinates.

```{r}
ara_mean_temp_2016_access1_rcp45_req_query <-
  ara_mean_temp_2016_access1_rcp45_req |>
  ccia_add_netcdf_subset_query(vars = c('tas_djf', 'tas_mam', 'tas_jja', 'tas_son'),
                          bbox = c(151, -33, 151.1, -32.9))

ara_mean_temp_2016_access1_rcp45_req_query
```

Then we perform the request and save the object using `ccia_perform_query()`. The return
value is the file path

```{r}
temp_file_nc <- tempfile(fileext = '.nc')

ara_mean_temp_2016_access1_rcp45_req_filepath <-
  ara_mean_temp_2016_access1_rcp45_req_query |>
  ccia_perform_query(destfile = temp_file_nc)
```

Using the `stars` package

```{r}
temp_file_nc |>
  stars::read_ncdf()
```

