---
title: "Explore application ready aggregated data - Single point"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explore application ready aggregated data - Single point}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(climateChangeInAustralia)

# 
library(stars)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(units)
```

## Application ready aggregated

See [introduction to application ready data](https://www.climatechangeinaustralia.gov.au/en/obtain-data/application-ready-data/).
The aggregated files contain Monthly, seasonal and annual data as time-series. They 
contain a single time value per range period (e.g. 2016-2045 has only 2031-01-01) per grid point (lat/lon),
and then model values for monthly, seasonal and annual data (the remaining 'columns')

### A small download

Note how the datetime_start have no effect ...

```{r}
ccia_dataset_urls <- ccia_dataset_urls()

dataset_url <- 
  ccia_dataset_urls$
  application_ready_aggregated$
  Mean_Temperature$
  `2016-2045`$
  `tas_aus_ACCESS1-0_rcp45_r1i1p1_CSIRO-MnCh-wrt-1986-2005-Scl_v1_mon_seasavg_2016-2045_clim.nc`

dataset_query <-
  dataset_url |>
  ccia_create_dataset_request_url(access_type = ccia_access_types$NetcdfSubset) |>
  ccia_add_netcdf_subset_query(vars = 'all',
                               lat = -33.86,
                               lon = 151.20,
                               # No effect (in this case)
                               datetime_start = '2016-01-01',
                               datetime_end = '2045-12-31',
                               datetime_step = 365)

dataset_download_filepath <-
  dataset_query |>
  ccia_perform_query(destfile = tempfile(fileext = 'tas_aus_ACCESS1-0_rcp45_seasavg_2016-2045.nc'))

structure(
  file.size(dataset_download_filepath),
  class = 'object_size'
) |>
  format(units = 'MB')
```

```{r}
ncdf_tibble <-
  dataset_download_filepath |>
  stars::read_ncdf() |>
  tibble::as_tibble()

ncdf_tibble
```

```{r}
# monthly, seasonal and annual data
names(ncdf_tibble) |>
  cat(sep = '\n')

# time range
range(ncdf_tibble$time)
```

### A time-series download

```{r}
ccia_dataset_urls <- ccia_dataset_urls()

# Extract model 'ACCESS1-0_rcp85' through element names
series_urls <-
  ccia_dataset_urls$application_ready_aggregated$Mean_Temperature |>
  lapply(function(urls){
    urls[grepl('ACCESS1-0_rcp85', names(urls))]
  })

# Expect 4 results
sapply(series_urls, names)

# Build requests
series_requests <-
  lapply(series_urls, function(url){
    
    url |>
      ccia_create_dataset_request_url(access_type = ccia_access_types$NetcdfSubset) |>
      ccia_add_netcdf_subset_query(vars = 'all',
                                   lat = -33.86,
                                   lon = 151.20)
    
  })

# Download data
series_filepaths <-
  lapply(series_requests, function(req){
    
    filename <- sub('.*\\d/(.*\\.nc).*', '\\1', req$url)
  
    req |>
      ccia_perform_query(destfile =  tempfile(pattern = 'ccia-', fileext = filename))
    
  })
```

Import into R and do things.

```{r}
# Merge data into a tibble
series_tibble <-
  purrr::map_df(series_filepaths, function(filepath){
    filepath |>
      stars::read_ncdf() |>
      tibble::as_tibble()
  })
```

```{r}
series_tibble

range(c(series_tibble$lon, series_tibble$lat))
range(series_tibble$time)
unique(series_tibble$time)
```

### Some data visualisation

#### Annual

```{r}
series_tibble |>
  ggplot(aes(x = time, y = tas_annual)) +
  geom_point() +
  labs(title = 'Annual mean temperature', 
       subtitle = 'at: -33.85, 151.2, model: ACCESS1-0_rcp85')
```

#### Monthly

```{r}
series_tibble |>
  dplyr::select(time, dplyr::contains(month.name)) |>
  tidyr::pivot_longer(-time) |>
  dplyr::arrange(time) |>
  dplyr::mutate(month = sub('tas_', '', name),
                year = format(time, '%Y'),
                month = forcats::fct_inorder(month)) |>
  ggplot(aes(x = month, y = value, shape = year, group = year)) +
  geom_line() +
  geom_point() +
  labs(title = 'Monthly mean temperature', 
       subtitle = 'at: -33.85, 151.2, model: ACCESS1-0_rcp85') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.33))
```

#### Seasonal

```{r}
series_tibble |>
  dplyr::select(time, tas_djf, tas_mam, tas_jja, tas_son) |>
  tidyr::pivot_longer(-time) |>
  dplyr::arrange(time) |>
  dplyr::mutate(season = dplyr::case_when(
    name == 'tas_djf' ~ 'Summer',
    name == 'tas_mam' ~ 'Autumn',
    name == 'tas_jja' ~ 'winter',
    name == 'tas_son' ~ 'Spring',
  ),
  year = format(time, '%Y'),
  season = forcats::fct_inorder(season)) |>
  ggplot(aes(x = season, y = value, shape = year, group = year)) +
  geom_line() +
  geom_point() +
  labs(title = 'Seasonal mean temperature', 
       subtitle = 'at: -33.85, 151.2, model: ACCESS1-0_rcp85') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.33))
```



