---
title: "Explore application ready aggregated data - Grid of points"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explore application ready aggregated data - Grid of points}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(climateChangeInAustralia)

# 
library(stars)
library(sf)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(units)
# 
library(boundariesNSW)
```

## Application ready aggregated

See [introduction to application ready data](https://www.climatechangeinaustralia.gov.au/en/obtain-data/application-ready-data/).
The aggregated files contain Monthly, seasonal and annual data as time-series. They 
contain a single time value per range period (e.g. 2016-2045 has only 2031-01-01) per grid point (lat/lon),
and then model values for monthly, seasonal and annual data (the remaining 'columns')

### A small download

Here we will download `Mean_Temperature` for `2016-2045` for `ACCESS1-0` and `rcp45`, and
then constrain to the the the Greater Sydney Area bounding box using `boundariesNSW`.

Get the boundary information.

```{r}
greater_sydney_boundary <- 
  bnsw_get_map('gccsa2016') |>
  dplyr::filter(GCC_NAME16 == 'Greater Sydney')

greater_sydney_boundary_bbox <- sf::st_bbox(greater_sydney_boundary)
greater_sydney_boundary_bbox
```

Download the data. Note how the datetime_start have no effect ...

```{r}
ccia_dataset_urls <- ccia_dataset_urls()

dataset_url <- 
  ccia_dataset_urls$
  application_ready_aggregated$
  Mean_Temperature$
  `2016-2045`$
  `tas_aus_ACCESS1-0_rcp45_r1i1p1_CSIRO-MnCh-wrt-1986-2005-Scl_v1_mon_seasavg_2016-2045_clim.nc`

dataset_query <-
  dataset_url |>
  ccia_create_dataset_request_url(access_type = ccia_access_types$NetcdfSubset) |>
  ccia_add_netcdf_subset_query(vars = 'all',
                               # Greater Sydney Area
                               bbox = greater_sydney_boundary_bbox,
                               # No effect (in this case)
                               datetime_start = '2016-01-01',
                               datetime_end = '2045-12-31',
                               datetime_step = 365)



dataset_download_filepath <-
  dataset_query |>
  ccia_perform_query(destfile = tempfile(fileext = 'tas_aus_ACCESS1-0_rcp45_seasavg_2016-2045.nc'))

structure(
  file.size(dataset_download_filepath),
  class = 'object_size'
) |>
  format(units = 'MB')
```

Import into R, wrangle and make a plot.

```{r}
tas_ncdf <-
  dataset_download_filepath |>
  stars::read_stars()

tas_ncdf
```

```{r}
tas_ncdf_month <-
  # first 12 values are monthly estimates
  tas_ncdf[1:12] |>
  # merge attributes (temperatures) into a new dimension
  # and set it's values to month names
  merge() |>
  st_set_dimensions(4, values = month.name) |>
  # pop the time dimension
  split('time') |>
  # rename
  st_set_dimensions(names = c('x', 'y', 'time')) |>
  setNames('Temp')

# Match CRS to boundary for cropping (as it's missing also)
tas_ncdf_month <-
  tas_ncdf_month |>
  st_set_crs(value = st_crs(greater_sydney_boundary))
```

```{r, fig.width=8, fig.height=8}
ggplot() +
  geom_stars(data = tas_ncdf_month[greater_sydney_boundary]) +
  facet_wrap(~time) +
  theme_void() +
  labs(title = 'Mean monthly temperature',
       subtitle = 'model: ACCESS1, scenario: rcp45')
```

### A time-series download

Needs a bit of R jujitsu ..

```{r}
ccia_dataset_urls <- ccia_dataset_urls()

# Extract model 'ACCESS1-0_rcp85' through element names
series_urls <-
  lapply(ccia_dataset_urls$application_ready_aggregated$Mean_Temperature, function(urls){
    
    urls[grepl('ACCESS1-0_rcp85', names(urls))]
    
  })

# Expect 4 results
sapply(series_urls, names)

# Build requests
series_requests <-
  lapply(series_urls, function(url){
    
    url |>
      ccia_create_dataset_request_url(access_type = ccia_access_types$NetcdfSubset) |>
      ccia_add_netcdf_subset_query(vars = 'all',
                                   # Greater Sydney Area
                                   bbox = greater_sydney_boundary_bbox)
  })

# Download data
series_filepaths <-
  lapply(series_requests, function(req){
    
    filename <- sub('.*\\d/(.*\\.nc).*', '\\1', req$url)
    
    req |>
      ccia_perform_query(destfile = tempfile(pattern = 'ccia-', fileext = filename))
    
  })
```

Import into R, wrangle and make some plota.

```{r}
series_stars <-
  lapply(series_filepaths, stars::read_ncdf, var = 'tas_annual')

do.call(c, series_stars) |>
  plot(key.pos = 1,
       box_col = 0)
```

#### Monthly

```{r}
month_var_names <- paste0('tas_', tolower(month.name))

series_stars <-
  lapply(series_filepaths, stars::read_ncdf, var = month_var_names)

wrangle_months <- function(x){
  
  year <-
    x |>
    st_get_dimension_values('time') |>
    format('%Y')
  
  x[month_var_names] |>
    merge() |>
    st_set_dimensions(4, values = paste(month.abb, year)) |>
    split('time') |>
    st_set_dimensions(names = c('x', 'y', 'time')) |>
    setNames('Temp')
}

series_stars_wrangled <-
  lapply(rev(series_stars), wrangle_months) |>
  do.call(what = c)

series_stars_wrangled <-
  series_stars_wrangled |>
  st_transform(crs = st_crs(greater_sydney_boundary)) 

series_stars_wrangled <- 
  series_stars_wrangled[greater_sydney_boundary]
```

```{r}
# downsampled for faster plotting
ggplot() +
  geom_stars(data = series_stars_wrangled, downsample = c(2,2,0)) +
  facet_wrap(~time, ncol = 4, dir = 'v') +
  theme_void()
```

#### Seasonal

```{r}
season_var_names <- 
  c(summer = 'tas_djf', 
    autumn = 'tas_mam', 
    winter = 'tas_jja',
    spring = 'tas_son')

series_stars <-
  lapply(series_filepaths, stars::read_ncdf, var = season_var_names)

wrangle_seasons <- function(x){
  
  year <-
    x |>
    st_get_dimension_values('time') |>
    format('%Y')
  
  x[season_var_names] |>
    merge() |>
    st_set_dimensions(4, values = paste(names(season_var_names), year)) |>
    split('time') |>
    st_set_dimensions(names = c('x', 'y', 'time')) |>
    setNames('Temp')
}

series_stars_wrangled <-
  lapply(rev(series_stars), wrangle_seasons) |>
  do.call(what = c)

series_stars_wrangled <-
  series_stars_wrangled |>
  st_transform(crs = st_crs(greater_sydney_boundary)) 

series_stars_wrangled <- 
  series_stars_wrangled[greater_sydney_boundary]

ggplot() +
  geom_stars(data = series_stars_wrangled, downsample = c(1,1,0)) +
  facet_wrap(~time, ncol = 4) +
  scale_fill_continuous(type = 'viridis') +
  theme_void()
```



